---
title: "Wind Indicator Experiments"
author: "laura gruenburg"
date: "8/14/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(ggplot2)
```

## Load the wind data


```{r cars}
setwd("/Users/nyelab/Desktop/NYB Indicators/NYB_Indicators_Calculations/Final_Timeseries_Figures")
b_nyh = read.csv("Wind_buoy2_Aug_14_2023.csv")
b_tt  = read.csv("Wind_buoy4_Aug_14_2023.csv")
b_isl = read.csv("Wind_buoy5_Aug_14_2023.csv")

```

## Get months

```{r}
b_nyh$month <- format(as.Date(b_nyh$datetime), "%m")
b_tt$month  <- format(as.Date(b_tt$datetime), "%m")
b_isl$month <- format(as.Date(b_isl$datetime), "%m")
```

## Exclude some really early data from Texas Tower

```{r}
b_tt <- b_tt[b_tt$year > 1985, ]
```

## Experimenting with winds greater than a certain threshhold (90th percentile)
## or pressure lower than a certain threshhold (10th percentile)

```{r}
pr_10percentile<- b_nyh %>% 
  group_by(month) %>% 
  summarise(quantile(air_pressure , probs=.1, na.rm = TRUE))

colnames(pr_10percentile)[2] ="low_pr"
Sub <- b_nyh[b_nyh$month == '01', ]
low <- Sub[Sub$air_pressure <= pr_10percentile[pr_10percentile$month == '01', ]$low_pr, ]

for (i in pr_10percentile$month){
  if (i =='01') next
  Sub <- b_nyh[b_nyh$month == i, ]
  low_sub <- Sub[Sub$air_pressure <= pr_10percentile[pr_10percentile$month == '01', ]$low_pr, ]
  low <- rbind(low,low_sub)
}

low$yr_month <- format(as.Date(low$datetime), "%m-%Y")

```

```{r}
gust_90percentile<- b_nyh %>% 
  group_by(month) %>% 
  summarise(quantile(gust , probs=.9, na.rm = TRUE))

colnames(gust_90percentile)[2] ="speed"
Subg <- b_nyh[b_nyh$month == i, ]
gusty <- Subg[Subg$gust >= gust_90percentile[gust_90percentile$month == '01', ]$speed, ]

for (i in gust_90percentile$month){
  if (i =='01') next
  Subg <- b_nyh[b_nyh$month == i, ]
  gust_sub <- Subg[Subg$gust >= gust_90percentile[gust_90percentile$month == '01', ]$speed, ]
  gusty <- rbind(gusty,gust_sub)
}

gusty$yr_month <- format(as.Date(gusty$datetime), "%m-%Y")
```

## Including Plots

```{r pressure, echo=FALSE}
## Temperature by month
ggplot(low, aes(x=year, y=air_temperature)) + 
  geom_point(size=.05)+facet_wrap(~month)

```

```{r pressure, echo=FALSE}
## Gusts in knots

ggplot(gusty, aes(x=year, y=gust*(1/1852)*(3600/1))) + 
  geom_point(size=.05)+facet_wrap(~month)

```
## Seasonal wind in knots - small craft advisory and tropical storm warning

```{r}

daily_wind_nyh <- b_nyh %>% group_by(year, yday) %>%
  summarize(wind = mean(wind_spd)*(1/1852)*(3600/1),
            dir = mean(wind_dir),
            m = mean(as.numeric(month)))

daily_wind_nyh$season = 'fall'
daily_wind_nyh[daily_wind_nyh$m == 1,]$season = 'winter'
daily_wind_nyh[daily_wind_nyh$m == 2,]$season = 'winter'
daily_wind_nyh[daily_wind_nyh$m == 3,]$season = 'winter'

daily_wind_nyh[daily_wind_nyh$m == 4,]$season = 'spring'
daily_wind_nyh[daily_wind_nyh$m == 5,]$season = 'spring'
daily_wind_nyh[daily_wind_nyh$m == 6,]$season = 'spring'

daily_wind_nyh[daily_wind_nyh$m == 7,]$season = 'summer'
daily_wind_nyh[daily_wind_nyh$m == 8,]$season = 'summer'
daily_wind_nyh[daily_wind_nyh$m == 9,]$season = 'summer'

daily_wind_nyh$category = 'No advisory'

daily_wind_nyh[!is.na(daily_wind_nyh$wind) & daily_wind_nyh$wind >= 18 & daily_wind_nyh$wind <=33, ]$category ='small craft warning'

daily_wind_nyh[!is.na(daily_wind_nyh$wind) & daily_wind_nyh$wind>= 34 & daily_wind_nyh$wind<=64, ]$category = 'tropical storm warning'

### 

daily_wind_isl <- b_isl %>% group_by(year, yday) %>%
  summarize(wind = mean(wind_spd)*(1/1852)*(3600/1),
            dir = mean(wind_dir),
            m = mean(as.numeric(month)))

daily_wind_isl$season = 'fall'
daily_wind_isl[daily_wind_isl$m == 1,]$season = 'winter'
daily_wind_isl[daily_wind_isl$m == 2,]$season = 'winter'
daily_wind_isl[daily_wind_isl$m == 3,]$season = 'winter'

daily_wind_isl[daily_wind_isl$m == 4,]$season = 'spring'
daily_wind_isl[daily_wind_isl$m == 5,]$season = 'spring'
daily_wind_isl[daily_wind_isl$m == 6,]$season = 'spring'

daily_wind_isl[daily_wind_isl$m == 7,]$season = 'summer'
daily_wind_isl[daily_wind_isl$m == 8,]$season = 'summer'
daily_wind_isl[daily_wind_isl$m == 9,]$season = 'summer'

daily_wind_isl$category = 'No advisory'

daily_wind_isl[!is.na(daily_wind_isl$wind) & daily_wind_isl$wind >= 18 & daily_wind_isl$wind <=33, ]$category ='small craft warning'

daily_wind_isl[!is.na(daily_wind_isl$wind) & daily_wind_isl$wind>= 34 & daily_wind_isl$wind<=64, ]$category = 'tropical storm warning'


###

daily_wind_tt <- b_tt %>% group_by(year, yday) %>%
  summarize(wind = mean(wind_spd)*(1/1852)*(3600/1),
            dir = mean(wind_dir),
            m = mean(as.numeric(month)))

daily_wind_tt$season = 'fall'
daily_wind_tt[daily_wind_tt$m == 1,]$season = 'winter'
daily_wind_tt[daily_wind_tt$m == 2,]$season = 'winter'
daily_wind_tt[daily_wind_tt$m == 3,]$season = 'winter'

daily_wind_tt[daily_wind_tt$m == 4,]$season = 'spring'
daily_wind_tt[daily_wind_tt$m == 5,]$season = 'spring'
daily_wind_tt[daily_wind_tt$m == 6,]$season = 'spring'

daily_wind_tt[daily_wind_tt$m == 7,]$season = 'summer'
daily_wind_tt[daily_wind_tt$m == 8,]$season = 'summer'
daily_wind_tt[daily_wind_tt$m == 9,]$season = 'summer'


daily_wind_tt$category = 'No advisory'

daily_wind_tt[!is.na(daily_wind_tt$wind) & daily_wind_tt$wind >= 18 & daily_wind_tt$wind <=33, ]$category ='small craft warning'

daily_wind_tt[!is.na(daily_wind_tt$wind) & daily_wind_tt$wind>= 34 & daily_wind_tt$wind<=64, ]$category = 'tropical storm warning'

```

``` {r}

warnings_nyh<- daily_wind_nyh %>% group_by(year, season) %>% 
  summarise(n_scw = sum(category == 'small craft warning'),
            n_tsw = sum(category == 'tropical storm warning'))

warnings_tt<- daily_wind_tt %>% group_by(year, season) %>% 
  summarise(n_scw = sum(category == 'small craft warning'),
            n_tsw = sum(category == 'tropical storm warning'))

warnings_isl<- daily_wind_isl %>% group_by(year, season) %>% 
  summarise(n_scw = sum(category == 'small craft warning'),
            n_tsw = sum(category == 'tropical storm warning'))

```

## Plots of warnings

```{r}
warnings_nyh$buoy <- 'New York Harbor'
warnings_tt$buoy <- 'Texas Tower'
warnings_isl$buoy <- 'South of Islip'

all_warnings <- rbind(warnings_nyh, warnings_tt, warnings_isl)


variable_names <- list(
  "winter" = "Winter" ,
  "spring" = "Spring",
  "summer" = "Summer",
  "fall" = "Fall"
)



buoy_names <- c('New York Harbor', 'South of Islip', 'Texas Tower')


variable_labeller2 <- function(variable,value){
  if (variable=='season') {
  return(variable_names[value])
  } else {
    return(buoy_names)
  }
}
ggplot(all_warnings, aes(x=year, y=n_scw, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)

ggplot(all_warnings, aes(x=year, y=n_tsw, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)

```
## Wind direction plots

```{r}

daily_wind_nyh$buoy <- 'New York Harbor'
daily_wind_tt$buoy <- 'Texas Tower'
daily_wind_isl$buoy <- 'South of Islip'

all_wind <- rbind(daily_wind_nyh, daily_wind_tt, daily_wind_isl)
all_wind$year_frac <- all_wind$year + (all_wind$yday/365)

variable_names <- list(
  "winter" = "Winter" ,
  "spring" = "Spring",
  "summer" = "Summer",
  "fall" = "Fall"
)



buoy_names <- c('New York Harbor', 'South of Islip', 'Texas Tower')


variable_labeller2 <- function(variable,value){
  if (variable=='season') {
  return(variable_names[value])
  } else {
    return(buoy_names)
  }
}

ggplot(all_wind[all_wind$year == 2009, ], aes(x=yday, y=dir, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)

ggwindrose(all_wind, wind,dir)
```

## Windrose plotting function

```{r}

# WindRose.R
require(ggplot2)
require(RColorBrewer)

plot.windrose <- function(data,
                      spd,
                      dir,
                      spdres = 2,
                      dirres = 30,
                      spdmin = 2,
                      spdmax = 20,
                      spdseq = NULL,
                      palette = "YlGnBu",
                      countmax = NA,
                      debug = 0){


# Look to see what data was passed in to the function
  if (is.numeric(spd) & is.numeric(dir)){
    # assume that we've been given vectors of the speed and direction vectors
    data <- data.frame(spd = spd,
                       dir = dir)
    spd = "spd"
    dir = "dir"
  } else if (exists("data")){
    # Assume that we've been given a data frame, and the name of the speed 
    # and direction columns. This is the format we want for later use.    
  }  

  # Tidy up input data ----
  n.in <- NROW(data)
  dnu <- (is.na(data[[spd]]) | is.na(data[[dir]]))
  data[[spd]][dnu] <- NA
  data[[dir]][dnu] <- NA

  # figure out the wind speed bins ----
  if (missing(spdseq)){
    spdseq <- seq(spdmin,spdmax,spdres)
  } else {
    if (debug >0){
      cat("Using custom speed bins \n")
    }
  }
  # get some information about the number of bins, etc.
  n.spd.seq <- length(spdseq)
  n.colors.in.range <- n.spd.seq - 1

  # create the color map
  spd.colors <- colorRampPalette(brewer.pal(min(max(3,
                                                    n.colors.in.range),
                                                min(9,
                                                    n.colors.in.range)),                                               
                                            palette))(n.colors.in.range)

  if (max(data[[spd]],na.rm = TRUE) > spdmax){    
    spd.breaks <- c(spdseq,
                    max(data[[spd]],na.rm = TRUE))
    spd.labels <- c(paste(c(spdseq[1:n.spd.seq-1]),
                          '-',
                          c(spdseq[2:n.spd.seq])),
                    paste(spdmax,
                          "-",
                          max(data[[spd]],na.rm = TRUE)))
    spd.colors <- c(spd.colors, "grey50")
  } else{
    spd.breaks <- spdseq
    spd.labels <- paste(c(spdseq[1:n.spd.seq-1]),
                        '-',
                        c(spdseq[2:n.spd.seq]))    
  }
  data$spd.binned <- cut(x = data[[spd]],
                         breaks = spd.breaks,
                         labels = spd.labels,
                         ordered_result = TRUE)
  # clean up the data
  data. <- na.omit(data)

  # figure out the wind direction bins
  dir.breaks <- c(-dirres/2,
                  seq(dirres/2, 360-dirres/2, by = dirres),
                  360+dirres/2)  
  dir.labels <- c(paste(360-dirres/2,"-",dirres/2),
                  paste(seq(dirres/2, 360-3*dirres/2, by = dirres),
                        "-",
                        seq(3*dirres/2, 360-dirres/2, by = dirres)),
                  paste(360-dirres/2,"-",dirres/2))
  # assign each wind direction to a bin
  dir.binned <- cut(data[[dir]],
                    breaks = dir.breaks,
                    ordered_result = TRUE)
  levels(dir.binned) <- dir.labels
  data$dir.binned <- dir.binned

  # Run debug if required ----
  if (debug>0){    
    cat(dir.breaks,"\n")
    cat(dir.labels,"\n")
    cat(levels(dir.binned),"\n")       
  }  

  # deal with change in ordering introduced somewhere around version 2.2
  if(packageVersion("ggplot2") > "2.2"){    
    cat("Hadley broke my code\n")
    data$spd.binned = with(data, factor(spd.binned, levels = rev(levels(spd.binned))))
    spd.colors = rev(spd.colors)
  }

  # create the plot ----
  p.windrose <- ggplot(data = data,
                       aes(x = dir.binned,
                           fill = spd.binned)) +
    geom_bar() + 
    scale_x_discrete(drop = FALSE,
                     labels = waiver()) +
    coord_polar(start = -((dirres/2)/360) * 2*pi) +
    scale_fill_manual(name = "Wind Speed (m/s)", 
                      values = spd.colors,
                      drop = FALSE) +
    theme(axis.title.x = element_blank())

  # adjust axes if required
  if (!is.na(countmax)){
    p.windrose <- p.windrose +
      ylim(c(0,countmax))
  }

  # print the plot
  print(p.windrose)  

  # return the handle to the wind rose
  return(p.windrose)
}

```

## Windrose plots 
```{r}

all_wind <- all_wind[!is.na(all_wind$dir),  ]
plot.windrose(spd = all_wind$wind,
                   dir = all_wind$dir,
              dirres = 10)

p.wr2 <- plot.windrose(data = all_wind,
              spd = "wind",
              dir = "dir",
              dirres = 10)
# now generate the faceting
p.wr3 <- p.wr2 + facet_wrap(~buoy,
                            ncol = 3)

p.wr4 <- p.wr2 + facet_grid(rows=vars(season), cols = vars(buoy))+ theme(axis.text.x = element_blank(),
          axis.title.x = element_blank())


buoy_names <- c('New York Harbor', 'South of Islip', 'Texas Tower')


variable_labeller2 <- function(variable,value){
  if (variable=='season') {
  return(variable_names[value])
  } else {
    return(buoy_names)
  }
}

plot.windrose(data = all_wind, spd = "wind", dir = "dir")+
  facet_grid(season~buoy, space="free_x",  labeller= variable_labeller2)
```
```{r}
plot.windrose(data = all_wind[all_wind$buoy == "New York Harbor", ], spd = "wind", dir = "dir", dirres = 10)+
  facet_grid(rows=vars(year), cols = vars(season))+ theme(axis.text.x = element_blank(),
          axis.title.x = element_blank())

```



```{r}
#north = 337.5 = 22.5
#northeast = 22.5 - 67.5
#east = 67.5 - 112.5
#southeast = 112.5 - 157.5
#south =  157.5 - 202.5
#southwest = 202.5 - 247.5
#west = 247.5 - 292.5
#northwest = 292.5 - 337.5

all_wind$dir_group = 'NaN'

all_wind[all_wind$dir >= 337.5, ]$dir_group <- 'north'
all_wind[all_wind$dir < 22.5, ]$dir_group <- 'north'

all_wind[all_wind$dir >= 22.5 & all_wind$dir <67.5, ]$dir_group <- 'northeast'
all_wind[all_wind$dir >= 67.5 & all_wind$dir <112.5, ]$dir_group <- 'east'
all_wind[all_wind$dir >= 112.5 & all_wind$dir <157.5, ]$dir_group <- 'southeast'
all_wind[all_wind$dir >= 157.5 & all_wind$dir <202.5, ]$dir_group <- 'south'
all_wind[all_wind$dir >= 202.5 & all_wind$dir <247.5, ]$dir_group <- 'southwest'
all_wind[all_wind$dir >= 247.5 & all_wind$dir <292.5, ]$dir_group <- 'west'
all_wind[all_wind$dir >= 292.5 & all_wind$dir <337.5, ]$dir_group <- 'northwest'


direction_wind <- all_wind %>% group_by(year, season, buoy) %>%
  summarize(n_days  = sum(dir_group == 'north'),
            ne_days  = sum(dir_group == 'northeast'),
            e_days  = sum(dir_group == 'east'),
            se_days  = sum(dir_group == 'southeast'),
            s_days  = sum(dir_group == 'south'),
            sw_days  = sum(dir_group == 'southwest'),
            w_days  = sum(dir_group == 'west'),
            nw_days  = sum(dir_group == 'northwest'))

strengthdirection_wind <- all_wind %>% group_by(year, season, buoy, dir_group) %>%
  summarize(n_strength  = mean(wind, na.rm = TRUE),
            ne_strength  = mean(wind, na.rm = TRUE),
            e_strength  = mean(wind, na.rm = TRUE),
            se_strength  = mean(wind, na.rm = TRUE),
            s_strength  = mean(wind, na.rm = TRUE),
            sw_strength  = mean(wind, na.rm = TRUE),
            w_strength  = mean(wind, na.rm = TRUE),
            nw_strength  = mean(wind, na.rm = TRUE))


buoy_names <- c('New York Harbor', 'South of Islip', 'Texas Tower')


variable_labeller2 <- function(variable,value){
  if (variable=='season') {
  return(variable_names[value])
  } else {
    return(buoy_names)
  }
}

ggplot(direction_wind, aes(x=year, y=s_days, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Southerly Winds')

ggplot(direction_wind, aes(x=year, y=sw_days, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Southwesterly Winds')

ggplot(direction_wind, aes(x=year, y=nw_days, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Northwesterly Winds')

ggplot(direction_wind, aes(x=year, y=w_days, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Westerly Winds')


ggplot(strengthdirection_wind[strengthdirection_wind$dir_group == 'south', ],
       aes(x=year, y=n_strength, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Southerly Winds')

ggplot(strengthdirection_wind[strengthdirection_wind$dir_group == 'southwest', ],
       aes(x=year, y=n_strength, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Southwesterly Winds')

ggplot(strengthdirection_wind[strengthdirection_wind$dir_group == 'northwest', ],
       aes(x=year, y=n_strength, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Northwesterly Winds')


ggplot(strengthdirection_wind[strengthdirection_wind$dir_group == 'west', ],
       aes(x=year, y=n_strength, color=buoy))+
  geom_line(stat='identity')+
  facet_grid(season~buoy, scales="free", space="free_x",  labeller= variable_labeller2)+ ggtitle('Westerly Winds')