### River discharge metrics

This function takes only a single input (the USGS gage # for a river of choice "in quotes") and will imports all available daily stream flow data for that site. Using this dataset, it will then calculate the following metrics. 

1 date of center timing of flow mass ($date_CT), units are day of water year
2 date of observed peak flow ($date_max_flow), units are day of water year
3 maximum flow, ($max_flow), units are cubic feet per second
4 number of days in the data for a water year ($n_days)
5 mean annual flow ($mean_flow), units are cubic feet per second
6 date of spring onset of flow ($date_spring_onset) units are day of water year
7 total annual discharge

ADDED AUG 12, center of fall flow and center of spring flow as per Hodgkins et al. 2003 which is mroe relevant to the East Coast than some of the other ones. 

see Cayan et al. 2001 for spring onset of flow alogrithm description. Technically, the spring onset metric identifies the day on which the cumulative departure from the mean daily flow is most negative. This corresponds with the transition from winter low-flow conditions to the beginning of runoff in snow dominated systems but not sure how well this works in East Coast Rivers so should not be used probably.


see Wenger et al. 2010 for center timing of flow mass (CFM) description. Technically, CFM is the date on which more than 50% of the annual flow has occured. 


####NEW Version incorporating metrics of Hodgkins et al. 2003
```{r}
library(lubridate)
library(waterData)
library(lfstat)
GetFlowMet<-function(gage_id){
  
  
  #gage_id<-"01358000"
  dat<-importDVs(gage_id, code="00060", stat="00003")
  #assign water year values and reformat
  dat$dates<-as.POSIXlt(dat$dates)
  #dat$wyear<-water_year(dat$dates,origin="usgs")
  #Here use DAY OF YEAR instead of water year
  
  
  dat$wyear<-year(dat$dates)
  
  #remove na vales
  dat<-subset(dat,val!="NA")
  #make a vector of years
  water.years<-unique(as.numeric(as.character(dat$wyear)))
  dat$wyear<-as.numeric(as.character(dat$wyear))
  #calculate metrics for first water year
  i=1
  w<-subset(dat,wyear==water.years[i])#get this year data
  w$wyear.day<-1:nrow(w)#assign water year day
  CT.df<-subset(w,w$wyear.day==1)# make dataframe
  CT.df<-CT.df[,c("wyear","staid")]
  CT.df$date_CT<-round(sum(w$wyear.day*w$val) /sum(w$val)) #compute CT of mass
  CT.df$date_max_flow<-subset(w,w$val==max(w$val))[1,6] #date of max flow
  CT.df$max_flow<-w[w$wyear.day == CT.df[1,"date_max_flow"],"val"]
  CT.df$n_days<-nrow(w) #how many days in water year
  CT.df$mean_flow<-mean(w$val) # mean flow
  w$dif.mean<-w$val-CT.df[1,"mean_flow"] # diff of each day from mean flow
  w$cum.dif<-cumsum(w$dif.mean) # cumulative sum of diff from mean
  CT.df$date_spring_onset<-subset(w,cum.dif==min(w$cum.dif))[,6]
  CT.df$sum_flow<-sum(w$val)
  CT.df# this is the data for year i
  
  ##spring date max from Ho
  w$yday<-yday(w$dates)
  #spring/Winter Jan 1 to May 31
  spring<-w[w$yday < 152,]
  fall<-w[w$yday > 275 & w$yday < 365,]
  
  #Fall Oct1 to Dec 31
  CT.df$spring_date_max<-subset(spring,spring$val==max(spring$val))[1,9]
  CT.df$fall_date_max<-subset(fall,fall$val==max(fall$val))[1,9]
  
  #Spring_center
  spring$seasonal_volume<-cumsum(spring$val)
  thresh<-tail(spring$seasonal_volume,1)/2
  spring$sBigz<-spring$seasonal_volume > thresh
  sBigz<-spring[spring$sBigz == TRUE,]
  CT.df$spring_center<-sBigz[1,9]
  #Fall_center
    
    #Fall_center
  fall$seasonal_volume<-cumsum(fall$val)
  thresh<-tail(fall$seasonal_volume,1)/2
  fall$sBigz<-fall$seasonal_volume > thresh
  sBigz<-fall[fall$sBigz == TRUE,]
  CT.df$fall_center<-sBigz[1,9]
  #Fall_center
  
  ### loop through to do it for all other water years
  for(i in 2:length(water.years)){
    w<-subset(dat,wyear==water.years[i])#get this year data
    w$wyear.day<-1:nrow(w)#assign water year day
    CT.df.n<-subset(w,w$wyear.day==1)# make dataframe
    CT.df.n<-CT.df.n[,c("wyear","staid")]
    CT.df.n$date_CT<-round(sum(w$wyear.day*w$val) /sum(w$val)) #compute CT of mass
    CT.df.n$date_max_flow<-subset(w,w$val==max(w$val))[1,6] #date of max flow
    CT.df.n$max_flow<-w[w$wyear.day == CT.df.n[1,"date_max_flow"],"val"]
    CT.df.n$n_days<-nrow(w) #how many days in water year
    CT.df.n$mean_flow<-mean(w$val, NA.rm = TRUE) # mean flow
    w$dif.mean<-w$val-CT.df.n[1,"mean_flow"] # diff of each day from mean flow
    w$cum.dif<-cumsum(w$dif.mean) # cumulative sum of diff from mean
    CT.df.n$date_spring_onset<-subset(w,cum.dif==min(w$cum.dif))[1,6]
    CT.df.n$sum_flow<-sum(w$val)
    
    
      ##spring date max from Ho
  w$yday<-yday(w$dates)
  #spring/Winter Jan 1 to May 31
  spring<-w[w$yday < 152,]
  fall<-w[w$yday > 275 & w$yday < 365,]
  
  #Fall Oct1 to Dec 31
  CT.df.n$spring_date_max<-subset(spring,spring$val==max(spring$val))[1,9]
  CT.df.n$fall_date_max<-subset(fall,fall$val==max(fall$val))[1,9]
  
#Spring_center
  spring$seasonal_volume<-cumsum(spring$val)
  thresh<-tail(spring$seasonal_volume,1)/2
  spring$sBigz<-spring$seasonal_volume > thresh
  sBigz<-spring[spring$sBigz == TRUE,]
  CT.df.n$spring_center<-sBigz[1,9]
  #Fall_center
    
    #Fall_center
  fall$seasonal_volume<-cumsum(fall$val)
  thresh<-tail(fall$seasonal_volume,1)/2
  fall$sBigz<-fall$seasonal_volume > thresh
  sBigz<-fall[fall$sBigz == TRUE,]
  CT.df.n$fall_center<-sBigz[1,9]
  #Fall_center
    
    
    CT.df<-rbind(CT.df,CT.df.n)
  }
  
  CT.df
}
```


Use the function to calculate and format NYB indicator.
```{r}
#packages lfstat and waterData must be loaded
library(lfstat)
library(waterData)

hudson<-"01358000"#Hudson River at green island
flowdat_hudson<-GetFlowMet(hudson)#Run the function
str(flowdat_hudson)#74 years of data

#important to get rid of data from years with incomplete data
flowdat_hudson<-subset(flowdat_hudson,flowdat_hudson$n_days >=365)# get rid of..
cfs_to_cms<-0.0283168#conversion to metric for flow volume
```



```{r}
plot(flowdat_hudson$wyear, flowdat_hudson$mean_flow, type = "l")
plot(flowdat_hudson$wyear, flowdat_hudson$date_max_flow, type = "b")

plot(flowdat_hudson$wyear, flowdat_hudson$spring_date_max, type = "b")
points(flowdat_hudson$wyear, flowdat_hudson$spring_center, type = "b", col = "green")

plot(flowdat_hudson$wyear, flowdat_hudson$fall_date_max, type = "b")
points(flowdat_hudson$wyear, flowdat_hudson$fall_center, type = "b", col = "green")

plot(flowdat_hudson$wyear, flowdat_hudson$fall_date_max, type = "b")
plot(flowdat_hudson$wyear, flowdat_hudson$max_flow, type = "l")
plot(flowdat_hudson$wyear, flowdat_hudson$date_CT, type = "l")
```

Make the plots pretty and convert to cubic meters per second
```{r}
library(ggplot2)
library(mgcv)
ggplot(flowdat_hudson, aes(x = wyear)) + geom_line(aes(y = mean_flow*cfs_to_cms), color='black') + geom_abline(intercept = -87285.42*cfs_to_cms, slope = 51.36*cfs_to_cms) + labs (y = "Mean Flow m3/s", x = 'Year', title = 'Hudson Mean Flow at Green Island') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))

ggplot(flowdat_hudson, aes(x = wyear, y =mean_flow*cfs_to_cms )) + geom_line(color='black') + geom_smooth(method = lm, se = FALSE, color = 'blue') + geom_smooth(method = "gam", data=flowdat_hudson, formula = y ~s(x, k=15), se = FALSE, color = 'red', linetype = 'twodash') + labs (y = "Mean Flow m3/s", x = 'Year', title = 'Hudson Mean Flow at Green Island') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))


ggplot(flowdat_hudson, aes(x = wyear, y =mean_flow*cfs_to_cms)) + geom_line(color='black') + geom_smooth(method = "gam", data=flowdat_hudson, formula = y ~s(x, k=15), se = FALSE) + labs (y = "Mean Flow m3/s", x = 'Year', title = 'Hudson Mean Flow at Green Island') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))


gam_mod<-gam(mean_flow*cfs_to_cms ~ s(wyear, k = 15), data = flowdat_hudson)
summary(gam_mod)#check out model
gam.check(gam_mod)

ggplot(flowdat_hudson, aes(x = wyear)) + geom_line(aes(y = max_flow*cfs_to_cms), color='black') + geom_line(aes(y = date_max_flow*5), color='blue') + 
  scale_y_continuous(
    
    # Features of the first axis
    name = "Max Flow m3/s",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*1/5, name="Max Flow Day of Year")
  ) + labs (x = 'Year', title = 'Hudson Max Flow at Green Island') + theme(title=element_text(size = 16, face = 'bold'), axis.title.y.right=element_text(size = 14,color='blue'), axis.text= element_text(color = 'black', size = 12))
  
ggplot(flowdat_hudson, aes(x = wyear)) + geom_line(aes(y = max_flow*cfs_to_cms), color='black') + geom_line(aes(y = date_max_flow*5), color='blue') + 
  scale_y_continuous(
    
    # Features of the first axis
    name = "Max Flow m3/s",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*1/5, name="Max Flow Day of Year")
  ) + labs (x = 'Year', title = 'Hudson Max Flow at Green Island') + theme(title=element_text(size = 16, face = 'bold'), axis.title.y.right=element_text(size = 14,color='blue'), axis.text= element_text(color = 'black', size = 12))  
  
ggplot(dat, aes(x = dates, y = val*cfs_to_cms)) + geom_line() + labs(y = 'Flow Rate m3/s', x = ' ', title = 'Hudson River Flow at Green Island')+ theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))

ggplot(flowdat_hudson, aes(x = wyear)) + geom_line(aes(y = mean_flow*cfs_to_cms), color='black') + labs (y = "Mean Flow m3/s", x = 'Year', title = 'Hudson Mean Flow at Green Island') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))

linearMod_SC<- lm(spring_center ~ wyear, data=flowdat_hudson)
#No Significant Linear Trend
gam_mod_SC<-gam(spring_center ~ s(wyear, k = 3), data = flowdat_hudson)
summary(gam_mod_SC)#check out model
gam.check(gam_mod_SC)

ggplot(flowdat_hudson, aes(x = wyear)) + geom_line(aes(y = spring_center), color='black') + labs (y = "Day of Year", x = 'Year', title = 'Spring Center Flow Volume') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))

ggplot(flowdat_hudson, aes(x = wyear, y =spring_center)) + geom_line(color='black') + geom_smooth(method = "gam", data=flowdat_hudson, formula = y ~s(x, k=3), se = FALSE, color = 'red', linetype = 'twodash') + labs (y = "Day of Year", x = 'Year', title = 'Spring Center Flow Volume') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))


linearMod_FC<- lm(fall_center ~ wyear, data=flowdat_hudson)
#No significant linear trend
gam_mod_FC<-gam(fall_center ~ s(wyear, k = 3), data = flowdat_hudson)
summary(gam_mod_FC)#check out model
gam.check(gam_mod_FC)
gam_mod.d <- Deriv(gam_mod, n = 200)

ggplot(flowdat_hudson, aes(x = wyear)) + geom_line(aes(y = fall_center), color='black') + labs (y = "Day of Year", x = 'Year', title = 'Fall Center Flow Volume') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))

ggplot(flowdat_hudson, aes(x = wyear, y =fall_center)) + geom_line(color='black') + geom_smooth(method = "gam", data=flowdat_hudson, formula = y ~s(x, k=3), se = FALSE, color = 'red', linetype = 'twodash') + labs (y = "Day of Year", x = 'Year', title = 'Fall Center Flow Volume') + theme(title=element_text(size = 16, face = 'bold'), axis.title=element_text(size = 14), axis.text= element_text(color = 'black', size = 12))
```

Convert to CMS and write to final data folder. 
```{r}
hud_flowMean<-data.frame(Year = flowdat_hudson$wyear,
                        Variable = "Hudson_meanflow",
                        Val = flowdat_hudson$mean_flow,
                        Loc = "NYB",
                        N = "365")
hud_flowMAX<-data.frame(Year = flowdat_hudson$wyear,
                        Variable = "Hudson_MAX",
                        Val = flowdat_hudson$max_flow,
                        Loc = "NYB",
                        N = "365")
hud_flowSpringCenter<-data.frame(Year = flowdat_hudson$wyear,
                        Variable = "Hudson_SpringCenter",
                        Val = flowdat_hudson$spring_center ,
                        Loc = "NYB",
                        N = "365")
hud_flowFallCenter<-data.frame(Year = flowdat_hudson$wyear,
                        Variable = "Hudson_FallCenter",
                        Val = flowdat_hudson$fall_center,
                        Loc = "NYB",
                        N = "365")


flow<-rbind(hud_flowMean, hud_flowMAX, hud_flowSpringCenter, hud_flowFallCenter)

setwd("~/Desktop/NYB Indicators/Final_timeseries")
write.csv(flow, "RiverFlow_AUG_12_2020.csv") #Don't forget to change the name each year!



###info about indicators
hud_flowFallCenter
```
